---
title: "Creating Interactive Maps with Leaflet"
author: "Ronja Seitz"
date: "2025-03-10"
output: 
  html_document:
    self_contained: true
    theme: default

#params:
#  point_data:
#    label: "Input points:"
#    value: "data/BIG_faeces_loc_202406.csv"
#    input: file
#  northing_column:
#    label: "Column with UTM Northing"
#    value: "utm_northing"
#    input: text
#  easting_column:
#    label: "Column with UTM Easting"
#    value: "utm_easting"
#    input: text
#  date_column:
#    label:  "Column with date of observation"
#    value: "date..yyyy.mm.dd."
#    input: text
#  sex_column:
#    label: "Column with reindeer sex"
#    value: "reindeer_sex"
#    input: text
#  age_column:
#    label: "Column with reindeer age"
#    value: "reindeer_age"
#    input: text
    
---
This R markdown script can be used to easily visualise reindeer location data with different levels of informations and effort in creating the map.
We create interactive maps with the package leaflet (https://doi.org/10.32614/CRAN.package.leaflet).

#### **Install required packages**
This is only necessary once.
```{r code to install packages}
# install.packages("sf")
# install.packages("leaflet")
# install.packages("dplyr")
```

#### **Load required packages**
```{r setup, message=FALSE, warning=FALSE}
library(sf) #work with location data like points 
library(leaflet) #create interactive maps
library(dplyr) #to clean data
```

#### **1. Create interactive maps with leaflet**
Because commonly used base maps like Open Street Map are not optimized for Svalbard we first add base maps from the Norsk Polar institute (https://geodata.npolar.no/) as WMTS (Web Map Tile Services) to the leaflet interface.
To make different layers accesible we also add the possibility to select which basemap to show.

```{r}
# Step 1. Define WMTS URLs
svalbard_basemap_wmts <- "https://geodata.npolar.no/arcgis/rest/services/Basisdata/NP_Basiskart_Svalbard_WMTS_3857/MapServer/WMTS/tile/1.0.0/Basisdata_NP_Basiskart_Svalbard_WMTS_3857/default/default028mm/{z}/{y}/{x}.jpgpng"

svalbard_satellite_wmts <- "https://geodata.npolar.no/arcgis/rest/services/Basisdata/NP_Satellitt_Svalbard_WMTS_3857/MapServer/WMTS/tile/1.0.0/Basisdata_NP_Satellitt_Svalbard_WMTS_3857/default/default028mm/{z}/{y}/{x}.jpgpng"

# Step 2. Create Leaflet map with multiple basemaps and overlays
topo_map <- leaflet() %>%
  setView(lng = 15, lat = 78, zoom = 5) %>%  # Center on Svalbard
  
  # Add basemap options
  addProviderTiles(providers$Esri.WorldImagery, group = "Esri Satellite") %>%
  addProviderTiles(providers$OpenStreetMap, group = "OpenStreetMap") %>%
  addProviderTiles(providers$CartoDB.Positron, group = "CartoDB Light") %>%
  
  # Add the Svalbard topogrpahic Basemap WMTS layer
  addWMSTiles(
    baseUrl = svalbard_basemap_wmts,
    layers = "Basisdata_NP_Basiskart_Svalbard_WMTS_3857",
    options = WMSTileOptions(format = "image/jpgpng", transparent = TRUE),
    group = "Svalbard Basemap"
  ) %>%
  
  # Add the Svalbard Satellite WMTS layer
  addWMSTiles(
    baseUrl = svalbard_satellite_wmts,
    layers = "Basisdata_NP_Satellitt_Svalbard_WMTS_3857",
    options = WMSTileOptions(format = "image/jpgpng", transparent = TRUE),
    group = "Svalbard Satellite"
  ) %>%
  
  # Add layer control to switch between base layers and overlays
  addLayersControl(
    baseGroups = c("Esri Satellite", "OpenStreetMap", "CartoDB Light"),
    overlayGroups = c("Svalbard Basemap", "Svalbard Satellite"),
    options = layersControlOptions(collapsed = FALSE)
  )

# Step 3: Plot the map, which was defined in the previous step
topo_map
```

Add Reindeer Location Points
```{r}
# Step 1: Read the CSV file with the GPS data into R
data <- read.csv(data/BIG_faeces_loc_202406.csv)

# Step 2: Clean up the UTM coordinates using parameter column names
valid_range <- list(
  easting = c(400000, 900000),
  northing = c(6500000, 9500000)
)

data_clean <- data %>%
  filter(
    !is.na(.data[[utm_easting]]) & 
    !is.na(.data[[utm_northing]]) & 
    .data[[utm_easting]] >= valid_range$easting[1] & 
    .data[[utm_easting]] <= valid_range$easting[2] &
    .data[[utm_northing]] >= valid_range$northing[1] & 
    .data[[utm_northing]] <= valid_range$northing[2]
  )

# Step 3: Create an sf object with UTM coordinates (EPSG:25833)
data_sf <- st_as_sf(data_clean, coords = c(utm_easting, utm_northing), crs = 25833)

# Step 4: Reproject the coordinates to EPSG:4326 for use in leaflet
data_projected <- st_transform(data_sf, crs = 4326)

# Step 5: Extract the longitude and latitude
lng <- st_coordinates(data_projected)[, 1]
lat <- st_coordinates(data_projected)[, 2]

# Step 6: Create a copy of the base map
reindeer_map <- topo_map

# Step 7: Add markers to the Leaflet map
for (i in 1:length(lng)) {
  reindeer_map <- reindeer_map %>%
    addCircleMarkers(
      lng = lng[i], 
      lat = lat[i], 
      radius = 2, 
      color = "white", 
      fill = TRUE, 
      fillColor = "grey", 
      fillOpacity = 0.7, 
      opacity = 1,
      popup = paste(
        "Infos:", 
        "\nDate:", data_projected[[date..yyyy.mm.dd.]][i], 
        "\nSex:", data_projected[[reindeer_sex]][i], 
        "\nAge:", data_projected[[reindeer_age]][i]
      )
    )
}

# Step 8: Plot the map
reindeer_map

# Create stylish labels with HTML/CSS
label_text <- lapply(1:nrow(data_projected), function(i) {
  htmltools::HTML(paste0(
    "<div style='font-family: Arial, sans-serif; padding: 6px;'>
       <div style='background-color: #2c3e50; color: white; padding: 4px 8px; 
                   border-radius: 3px 3px 0 0; font-weight: bold;'>
         Observation ", i, "
       </div>
       <div style='padding: 6px; border: 1px solid #ddd; border-top: none;
                   border-radius: 0 0 3px 3px; background-color: #f9f9f9;'>
         <b>Date: </b>", data_projected[[date..yyyy.mm.dd.]][i], "<br/>
         <b>Sex: </b>", data_projected[[reindeer_sex]][i], "<br/>
         <b>Age: </b>", data_projected[[reindeer_age]][i], "<br/>
         <b>Coordinates:</b><br/>
         <span style='font-size: 0.9em; color: #555;'>
           ", round(lat[i], 4), "째N, ", round(lng[i], 4), "째E
         </span>
       </div>
     </div>"
  ))
})

# Define fixed colors for each sex
sex_colors <- c(
  "male" = "green",
  "female" = "yellow",
  "unknown" = "#95a5a6"
)

# Create advanced map with sex-based coloring
advanced_reindeer_markers_map <- topo_map

for (i in 1:length(lng)) {
  current_sex <- data_projected[[reindeer_sex]][i]
  marker_color <- ifelse(current_sex %in% names(sex_colors),
                        sex_colors[current_sex],
                        "#95a5a6")
  
  advanced_reindeer_markers_map <- advanced_reindeer_markers_map %>%  
    addCircleMarkers(  
      lng = lng[i],  
      lat = lat[i],  
      radius = 5,  
      color = "white",          
      fillColor = marker_color,
      fillOpacity = 0.8,  
      weight = 1.5,  
      popup = label_text[[i]],  
      label = paste("Observation", i)  
    )  
}

# Add legend for sex colors
advanced_reindeer_markers_map <- advanced_reindeer_markers_map %>%
  addLegend(
    position = "bottomright",
    colors = sex_colors,
    labels = names(sex_colors),
    title = "Reindeer Sex",
    opacity = 0.8
  )

# Output the map
advanced_reindeer_markers_map
```

```{r}
# Create stylish labels with HTML/CSS
label_text <- lapply(1:nrow(data_projected), function(i) {
  htmltools::HTML(paste0(
    "<div style='font-family: Arial, sans-serif; padding: 6px;'>
       <div style='background-color: #2c3e50; color: white; padding: 4px 8px; 
                   border-radius: 3px 3px 0 0; font-weight: bold;'>
         Observation ", i, "
       </div>
       <div style='padding: 6px; border: 1px solid #ddd; border-top: none;
                   border-radius: 0 0 3px 3px; background-color: #f9f9f9;'>
         <b>Date: </b>", data_projected[[date..yyyy.mm.dd.]][i], "<br/>
         <b>Sex: </b>", data_projected[[reindeer_sex]][i], "<br/>
         <b>Age: </b>", data_projected[[reindeer_age]][i], "<br/>
         <b>Coordinates:</b><br/>
         <span style='font-size: 0.9em; color: #555;'>
           ", round(lat[i], 4), "째N, ", round(lng[i], 4), "째E
         </span>
       </div>
     </div>"
  ))
})

# Define fixed colors for each sex
sex_colors <- c(
  "male" = "green",
  "female" = "yellow",
  "unknown" = "#95a5a6"  # Fixed spelling from "unkown" to "unknown"
)

# Create advanced map with sex-based coloring
advanced_reindeer_markers_map <- topo_map

for (i in 1:length(lng)) {
  current_sex <- data_projected[[reindeer_sex]][i]  # Fixed to use params
  marker_color <- ifelse(current_sex %in% names(sex_colors),
                        sex_colors[current_sex],
                        "#95a5a6")
  
  advanced_reindeer_markers_map <- advanced_reindeer_markers_map %>%  
    addCircleMarkers(  
      lng = lng[i],  
      lat = lat[i],  
      radius = 5,  
      color = "white",          
      fillColor = marker_color,
      fillOpacity = 0.8,  
      weight = 1.5,  
      popup = label_text[[i]],  
      label = paste("Observation", i)  
    )  
}

# Add legend for sex colors
advanced_reindeer_markers_map <- advanced_reindeer_markers_map %>%
  addLegend(
    position = "bottomright",
    colors = sex_colors,
    labels = names(sex_colors),
    title = "Reindeer Sex",
    opacity = 0.8
  )

# Output the map
advanced_reindeer_markers_map
```

